---
title: "MATH 216 Data Science Final Project (Fall 2016)"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
output: 
  html_document: 
    keep_md: yes
---


* Name: Yuchen Zhu
* Project title: Segregated Public Schools in New York City 
* Write-up on RPubs: <http://rpubs.com/ZhuYuchen/final>
* Abstract: 62 years have past since Brown v. Board of Education, a landmark US Supreme Court case that ruled state laws establishing separate public schools for black and white students to be unconstitutional. Although racial segregation by law at schools became obsolete ever since, racial segregation for a variety of other reasons persists until today.

```{r, echo=FALSE, message=FALSE}
#
# Include code for one plot here
#
library(ggplot2)
library(dplyr)
library(readxl)
library(tidyr)

racial_comp <- read_excel("racial_comp.xlsx", 
    sheet = "School")

baseline <- racial_comp %>% 
  group_by(year) %>% 
  summarise(num_schools = length(name),
            total = sum(total), 
            white = sum(num_white), prop_white = round(sum(num_white)/sum(total), 2),
            asian = sum(num_asian), prop_asian = round(sum(num_asian)/sum(total), 2),
            black = sum(num_black), prop_black = round(sum(num_black)/sum(total), 2),
            hispanic = sum(num_hispanic), prop_hispanic = round(sum(num_hispanic)/sum(total), 2),
            poverty = sum(num_poverty), prop_poverty = round(sum(num_poverty)/sum(total), 2)) %>%
  filter(year == "2015-16" | year == "2014-15")

baseline_tidy <- baseline %>% 
  gather(key = race, value = number, `white`, `asian`, `black`, `hispanic`) %>% 
  select(year, num_schools, total, race, number) %>% 
  mutate(prop = number/total) %>% 
  filter(year == "2015-16")
  
ggplot(data = baseline_tidy, aes(x = race, y = prop)) +
  geom_bar(stat = "identity") +
  labs(title = "Racial Composition of Public School Student, 2015-16")
```